\documentclass[a4paper,10pt]{article}


% Title Page
\title{Network Programming\\
\small{Assignment 3: Client-Server programming}}
\author{%
        \mbox{}\\
        Thomas Veerman\\
        \texttt{tveerman@cs.vu.nl}\\
        1329545\\
        \mbox{}\\
        Martijn Vermaat\\
        \texttt{mvermaat@cs.vu.nl}\\
        1362917
}


\begin{document}
\maketitle

\section{A Paper Storage Server}
\paragraph{Question 1}
It is possible to send any number of paper detail structures by using
an array:

\begin{tabbing}
\hspace{20pt}\=\kill
 \>\texttt{typedef document document\_list<>;}
\end{tabbing}

However, by omitting the size of the array \textsc{Sun RPC} actually reads that
as $2^{32} - 1$, so in fact it is limited. We have chosen to make a
linked list:

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{type}\=\texttt{def} \texttt{struct document\_node *document\_list;}\\ 
 \> \texttt{struct document\_node }\{ \+ \\
 \> \texttt{document item;}\\ 
 \> \texttt{document\_list next;} \- \\
 \> \} 
\end{tabbing}

Of course, in practice the linked list will be just as limited in size
on most systems. Both approaches would suffice for our paper storage
needs.

In any case, the \textsc{Sun RPC} code will make sure all papers will be
delivered at the caller as one data structure, even if it has to send
it in pieces.


\paragraph{Question 2}
The main problem when dealing with variable sized data and \textsc{Sun RPC} is
keeping memory usage low. Variable sized data is stored by dynamically
allocating memory and this needs to be freed sometime.

On the client it is all just a matter of
allocating a piece of memory large enough for the variable sized
data. We terminate soon after the RPC call is done, so this memory
will not be held allocated unnecessary long.

On the server however, we need to make sure memory consumption does
not grow unnecessary at each RPC call. All we can do is free the
return value of call $n$ as soon as we receive call $n+1$.
If the entire return value is static, we don't need to do anything,
because the same memory area will be used for call $n+1$.
However, with variable sized return values, we definitely want to
free any dynamically allocated regions as soon as possible (consider
fetching a 45 megabyte paper). \textsc{Sun RPC} provides functionality to
recursively free an entire data structure. The downside to using this,
is that we need to make sure the entire datastructure may be destroyed
(i.e. does not contain pointers to allocated regions we need later
on).

Another problem is that our paper data cannot be sent with a NULL byte
as terminator (the data might contain some itself). Therefore, the
length of the data needs to be sent with it. With an \textsc{XDR}
variable-length field, \textsc{Sun RPC} does just that.

This is the datatype we used for paper content. Omitting the length
implicitely defines it as $2^{32} - 1$:

\begin{tabbing}
\hspace{20pt}\=\kill
 \>\texttt{typedef opaque data <>;}
\end{tabbing}

This seems to be large enough for transmitting \textsc{PDF} files (we haven't
tested the \textsc{DOC} format).


\section{A Hotel Reservation Server}
\paragraph{Question 3}
Each method parameter should be passed by value, never by reference
(if you do try to send by reference everything related to that
particular object will be serialized first which can result in
megabytes of data if you are not careful with internal
references). This is because the variables have to be local if the
server wants to manipulate them; it cannot manipulate variables that
still reside at the client.

In our solution we send only integers (\textsc{Java} primitive, always used by
value) and Strings (immutable object, but serialization is needed) to
the server. This is the only data the server really needs to be able
to do its job, there is no need to send larger data structures, so we
do not risk sending too much data.

As result our server sends back a Set of Strings or a set of
Availability objects which in turn consist of 2 integers and a float
(all \textsc{Java} primitives), so also here we do not risk sending too much
data.

See also section \ref{sec:hotel} for a description of the server
interface.

\section{A Hotel Reservation Gateway}
\paragraph{Question 4}
We have used one thread per child as server structure. Though it is
unlikely that we will be sending big quantities of data, it can occur
that a client is on a very bad connection with high latency and low
throughput, which will slow down the server. Also, as we found out
first hand, clients that do not follow the protocol may keep the
connection open so that other clients cannot connect. By spawning a
thread for each incoming connection, this will not be a problem
anymore. Of course the server (not the gateway) must support
concurrent connections. To this end we made the book method on Room
objects synchronized, since that is the only place that might cause
problems when multiple clients want to book the same room.

\paragraph{Question 5}
To design a protocol that is suitable for multiple platforms and
languages, you need to make sure to keep the differences between
platforms and languages in mind. For example, where one platform
considers an int as a 16-bit value, others consider it a 32-bit
value. In your protocol you have to standardize these differences.

Our solution is to make a plain-text protocol with \textsc{ASCII}
character
encoding, \textsc{HTTP}-like. This is easy to debug, and virtually every
platform knows how to interpret \textsc{ASCII} character encoding. Almost all
data are strings, except for the price of a room, which is a float,
and the number of available rooms, which is an integer. We
used the standard \textsc{ASCII} encoding for a float (like the
\texttt{java.lang.Float.toString(float f)} function
returns). Integers are represented by their obvious \textsc{ASCII}
representation.

(For a more detailed description of our protocol, see the program
documentation for the hotel gateway in section \ref{sec:gateway}.)

\paragraph{Question 6}
We give a list of files each user needs for compiling the user
program.

To compile the \texttt{hotelserver} program:
\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{HotelServer.java} \\
 \> \texttt{SimpleHotel\_Skel.class \textrm{(Generated with \texttt{rmic} from \texttt{SimpleHotel.class})}} \\
 \> \texttt{Hotel.java} \\
 \> \texttt{SimpleHotel.java} \\
 \> \texttt{Room.java} \\
 \> \texttt{RoomType.java} \\
 \> \texttt{Availability.java} \\
 \> \texttt{NotBookedException.java} \\
 \> \texttt{NotAvailableException.java}
\end{tabbing}

To compile the \texttt{hotelgw} program:
\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{HotelGateway.java} \\
 \> \texttt{SimpleHotel\_Stub.class \textrm{(Generated with \texttt{rmic} from \texttt{SimpleHotel.class})}} \\
 \> \texttt{Hotel.java} \\
 \> \texttt{Availability.java} \\
 \> \texttt{NotAvailableException.java}
\end{tabbing}

To compile the \texttt{hotelclient} program:
\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{HotelClient.java} \\
 \> \texttt{SimpleHotel\_Stub.class \textrm{(Generated with \texttt{rmic} from \texttt{SimpleHotel.class})}} \\
 \> \texttt{Hotel.java} \\
 \> \texttt{Availability.java} \\
 \> \texttt{NotAvailableException.java}
\end{tabbing}

And to compile \texttt{SimpleHotel.class} we need:
\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{SimpleHotel.java} \\
 \> \texttt{Hotel.java} \\
 \> \texttt{Room.java} \\
 \> \texttt{RoomType.java} \\
 \> \texttt{Availability.java} \\
 \> \texttt{NotAvailableException.java} \\
 \> \texttt{NotBookedException.java}
\end{tabbing}

Note: This can be simplified a lot for \textsc{J2SE 5.0} and higher,
where the stub classes are generated dynamically and the \texttt{rmic}
compiler is not needed anymore. Our code uses other constructs
introduced in \textsc{J2SE 5.0}, so it will not compile on earlier
platforms anyway, but our \texttt{Makefile} does adhere to the
traditional dependencies described above.

For the skeleton classes a similar argument applies, but they are not
needed even in earlier versions of \textsc{Java}. From \textsc{J2SE 5.0}
on, they are not generated by \texttt{rmic} (by default). They are
ignored in our \texttt{Makefile}.


\section{Program documentation}
\subsection{A Paper Storage Server}

For the paper storage server we have written 2 programs:
\begin{center}
% use packages: array
\begin{tabular}{ l | p{9cm} }
paperclient.c & a program that connects to the paper server to store papers and retrieve paper info\\ \hline
paperserver.c & an iterative server that accepts papers and provides information about the papers\\
\end{tabular}
\end{center}

In the specification file \texttt{paperstorage.x} our procedures and data types are defined.

\paragraph{paperclient.c}
The program starts by typing \texttt{./paperclient} followed by a number of
parameters. How to use the program is printed by omitting the
parameters. When the program starts it parses the parameters and
executes the corresponding routine, otherwise it will print an error
message.

The \texttt{add}, \texttt{detail}, \texttt{fetch}, and list routines roughly work as follows. A
client structure is created, parameters are set, remote procedure is
called, and results are printed on standard output. That means that
also the fetched paper will be printed to standard out. You will have
to redirect the output if you want to save the file. We have chosen
not to save the file in, for example, the local directory because
stored papers might have the same filename.

The paper client uses the following parameters:

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{add <hostname> <author> <title> <filename.}{\texttt{pdf$\vert$doc}}\texttt{>} \\
 \> to add a paper to the paperstorage.
\end{tabbing}

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{paperclient detail <hostname> <number>} \\
 \> to retrieve details about a specific paper.
\end{tabbing}

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{paperclient fetch <hostname> <number>} \\
 \> to retrieve a stored paper.
\end{tabbing}

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{paperclient list <hostname>} \\
 \> to list the stored papers in paperstorage.
\end{tabbing}

\paragraph{paperserver.c}
The program starts by typing \texttt{./paperserver}. When the program starts it
waits for incoming connections (not actually part of the paperserver
object file), and handles the requests. It is an iterative server,
dictated by the nature of \textsc{Sun RPC}.

The server keeps an array of pointers to \texttt{paper\_t} structs. If the
array becomes full, it is resized. When a paper is added a piece of
memory is allocated and paper details and paper file are copied. Then
the number under which the paper is stored is sent back to the
client.

When a list request comes in a linked list containing information
about each paper is constructed and sent back.

The \texttt{detail} and \texttt{fetch} routines of the client are handled by the same
procedure on the server. The only difference is whether they use the
\emph{sparse} or \emph{detailed} flag. In both cases the author and
paper title is sent back, but in the case of fetch also the paper file
is copied into the result.

The server uses TCP connections, because the server routines would
have to be idempotent if we were to use UDP. This is no problem with
the \texttt{detail}, \texttt{fetch}, and \texttt{list} routines, however, we do not want to have
a paper stored multiple times with the add routine because of
connection problems.

Result values are static, because they must not be destroyed when our
routines exit (the \textsc{Sun RPC} code still has to send the result
to the calling program). However, we do like to free the return values
sometime. The earliest time we can do this is on the next remote
procedure call.
We choose to just free the result of a procedure on the next invocation
of that particular procedure, although with a bit of extra work we could
have made it so that result values are freed on the next invocation of
\emph{any} procedure.
We use the \textsc{Sun RPC}-provided \texttt{xdr\_free()} function to
free the memory of previous return values. To use it successfully
though, we had to make sure that there is no data in the freed
structure that we might need later on.

\paragraph{paperstorage.x}
As mentioned earlier it is possible to use different techniques to
transfer variably sized data. While it is possible to use an array
with (implicit) maximum size, we used a linked list as a datastructure
for any number of papers.

To make the interface a bit more elegant, we tried to use some generic
structures as much as posibble, sometimes with optional values. Also,
some procedures return more information than is currently used by our
client implementation, but we tried to choose the most elegant
solution.

When a procedure fails, we wanted to send back information telling why
something went wrong. Instead of incorporating every bit of data in a
structure, we used the more efficient unions.


\subsection{A Hotel Reservation Server}\label{sec:hotel}
Our implementation is tested on \textsc{GNU/Linux} using
\textsc{JDK 1.5.0\_06-b5} and \textsc{JDK 1.5.0\_09-b03}.

For the hotel reservation server we have written 9 classes:
\begin{center}
% use packages: array
\begin{tabular}{ l | p{7.0cm} }
Availability.java & structure holding price and the number of still available rooms for a specific room type\\ \hline
HotelClient.java & client program that connects to the hotel reservation server to book rooms and retrieve information\\ \hline
Hotel.java & interface for the hotel service running at the hotel reservation server\\ \hline
HotelServer.java & the Hotel Reservation Server\\ \hline
NotAvailableException.java & exception that occurs when a wanted room is not available\\ \hline
NotBookedException.java & exception handling an internal error\\ \hline
Room.java & a specific room that hold guest information when it is booked\\ \hline
RoomType.java & structure that holds data about the specific types of rooms\\ \hline
SimpleHotel.java & implementation of the hotel service\\
\end{tabular}
\end{center}

\paragraph{Availability.java}
This class is used in a \texttt{Set} to send information about which types of
rooms still have rooms left, and at which price.

\paragraph{HotelClient.java}
The program starts by typing \texttt{./hotelclient} (which is a shell
script wrapper around the actual \textsc{Java} program) followed by a
number of parameters. How to use the program is printed by omitting
the parameters.

We have designed \texttt{HotelClient} such that it can run on its own, but also
that it can be incorporated in a larger software package that needs
the book, details, and list functions. For example, to make a
graphical wrapper around it.

The \texttt{HotelClient} uses the following parameters:

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{hotelclient list <hostname>} \\
 \> to list still available rooms.
\end{tabbing}

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{hotelclient book <hostname> [type] <guest>} \\
 \> to book a room. If type (of room) is omitted a room will be chosen by the server.
\end{tabbing}

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{hotelclient guests <hostname>} \\
 \> to retrieve a list of guests that have booked a room.
\end{tabbing}

When the client receives an answer it is printed according to the following format:

\begin{center}
% use packages: array
\begin{tabular}{ l | p{9.3cm} }
\textbf{command} & \textbf{format}\\ \hline
list & ``\%3d room(s) of type \%s at \%.2f euros per night\textbackslash n'', for example, ``5 room(s) of type 1 at 150.00 euros per night\textbackslash n''\\ \hline
book & ``None'' or ``No rooms of type \emph{type} available'' when
there are no rooms available of a specific type, or ``Unfortunately we're full'' when type is omitted and every room is booked \\ \hline
guests & ``\emph{guest}\textbackslash n'', for example, ``Guillaume Pierre\textbackslash nThomas Veerman\textbackslash nMartijn Vermaat\textbackslash n'' \\
\end{tabular}
\end{center}

\paragraph{Hotel.java}
This is the interface for the hotel service running at the hotel reservation server. This interface is needed because it is obligatory to extend the java.rmi.Remote interface. The interface is as follows:

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{publ}\=\texttt{ic void bookRoom(int roomType, String guest)} \+ \\
 \> \texttt{throws NotAvailableException, RemoteException;} \- \\
 \> To book a room with a specific type.
\end{tabbing}

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{publ}\=\texttt{ic void bookRoom(String guest)} \+\\
 \> \texttt{throws NotAvailableException, RemoteException;} \- \\
 \> To book a room (a type will be arbitrarily chosen by the server).
\end{tabbing}

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{publ}\=\texttt{ic Set<String> registeredGuests()} \+\\
 \> \texttt{throws RemoteException;} \- \\
 \> To to retrieve a list of guests.
\end{tabbing}

\begin{tabbing}
\hspace{20pt}\=\kill
 \> \texttt{publ}\=\texttt{ic Set<Availability> availableRooms()} \+\\
 \> \texttt{throws RemoteException;} \- \\
 \> To retrieve a listing of available rooms and their types and prices.
\end{tabbing}


\paragraph{HotelServer.java}
The program starts by typing \texttt{./hotelserver} (which is a shell
script wrapper around the actual \textsc{Java} program). Before
starting the server the \texttt{rmiregistry} should already be
running. However, it is also possible to start your own registry using
the \texttt{java.rmi.registry.LocateRegistry} and
\texttt{java.rmi.registry.Registry} classes.

The internal workings are pretty simple. The registry is started, a
\texttt{SimpleHotel} (implementation of the \texttt{Hotel} interface) is instantiated
and bound in the registry. The registry and \texttt{SimpleHotel} take care of
the rest.

\paragraph{NotAvailableException.java}
An exception indicating that either the specified room type was not
available anymore, or that the hotel is full in the event the client
omits sending a specific room type.

\paragraph{NotBookedException.java}
An internal exception that occurs when we try to retrieve the name of a guest of a booked room that is not booked (thus there is no guest name). The client will never notice that this exception occurred.

\paragraph{Room.java}
Structure representing a room. Holds a guest's name when it is
booked. The \texttt{void book(String guest)} method is declared
\texttt{synchronized}, to prevent concurrency problems when the seem
room is tried to be booked twice at the same time.

\paragraph{RoomType.java}
Structure representing a room type. Holds a \texttt{int} representing the type
and a \texttt{float} representing price. Of course, the hotel server can attach
names to the numerical types. We have chosen to use \texttt{int}s because it is
easier to implement.

\paragraph{SimpleHotel.java}
Implementation of the \texttt{Hotel} interface. It generates a couple of room
types and then takes care of the \texttt{book}, \texttt{list}, and \texttt{guests} methods. The
\texttt{SimpleHotel} could easily be extended to store the information in a
database or file.

Although none of its methods are declared \texttt{synchronized}, a
\texttt{SimpleHotel} instance can be used concurrently. The source
code contains a few comments on this.


\subsection{A Hotel Reservation Gateway}\label{sec:gateway}

Our implementation is tested on \textsc{GNU/Linux} using
\textsc{JDK 1.5.0\_06-b5} and \textsc{JDK 1.5.0\_09-b03}.

For the paper storage server we have written 2 programs:
\begin{center}
% use packages: array
\begin{tabular}{ l | p{8.2cm} }
HotelGateway.java & a program that accepts incoming socket connections and translates them to the hotel reservation server via RMI to book rooms and retrieve information\\ \hline
hotelgwclient.c & a client program that connects to the gateway via sockets to book rooms and retrieve information\\
\end{tabular}
\end{center}

\paragraph{HotelGateway.java}
The program starts by typing \texttt{./hotelgw} (which is a shell
script wrapper around the actual \textsc{Java} program). When the
program starts it waits for incoming socket connections. Upon
accepting a socket connection, a thread is spawned that handles the
request further. In the mean time the server goes back to waiting for
new incoming connections.

The thread processes incoming data according to the protocol we
specify in more detail below. When at some point the client fails to follow the protocol
an error message is sent back giving some information that indicates
what went wrong. When everything is well the thread connects to the
Hotel Server using \textsc{RMI}, and invokes the corresponding methods just
like the HotelClient.

The Hotel Gateway assumes it is running on the same host as the Hotel
Server and therefore connects to the local RMI Registry. The gateway
does not need parameters.

\paragraph{hotelgwclient.c}
The program starts by typing \texttt{./hotelgwclient} followed by a
number of parameters. How to use the program is printed by omitting
the parameters. When the program starts it parses the parameters and
executes the corresponding routine, otherwise it will print an error
message. The parameters that can be passed are exactly the same as for
\texttt{HotelClient.java}.

Each of the routines (being ``list'', ``book'', and ``guests'')
connects to the gateway, sends the proper procedure to the gateway
according to the protocol, and reads the response. If something went
wrong (e.g. sending an malformed request or internal error on the
server side), the error message is printed. Otherwise the response is
parsed and printed using the following formats:

\begin{center}
% use packages: array
\begin{tabular}{ l | p{9.3cm} }
\textbf{command} & \textbf{format}\\ \hline
list & ``\%3d room(s) of type \%s at \%.2f euros per night\textbackslash n'', for example, ``5 room(s) of type 1 at 150.00 euros per night\textbackslash n''\\ \hline
book & None. It succeeds or an error message is printed \\ \hline
guests & ``\%s\textbackslash n'', for example, ``Guillaume Pierre\textbackslash nThomas Veerman\textbackslash nMartijn Vermaat\textbackslash n'' \\
\end{tabular}
\end{center}

The client does not need to do host to network conversions, since all data is interpreted byte for byte. Therefore we do not need to concern about byte order.

\paragraph{protocol}
As pointed out in our answer to question 5, we have decided to use a
plain-text protocol for communication over the sockets between the
gateway and clients.

A client sends a request message through the socket to the server. The
server responds with a response message.

The messages are defined as follows:

\begin{tabbing}
\hspace{20pt}\=\kill
 \> a req\=uest is a sequence of \textsc{ASCII} characters: \+\\
 \> request := procedure newline parameters newline\\
 \> procedure := `list' $\vert$ `book' $\vert$ `guests'\\
 \> parameters := empty $\vert$ ( type newline )? guest newline\\
 \> type := an ASCII string without newline or space characters\\
 \> guest := an ASCII string without newline\\
 \> newline := '\textbackslash n' \-\\
 \> a response is a sequence of \textsc{ASCII} characters:\+ \\
 \> response := status space description newline result newline \\
 \> status := `0' $\vert$ `1' $\vert$ `2'\\
 \> space := ` '\\
 \> description := an ASCII string without newline characters\\
 \> result := ( guest newline )* $\vert$ ( availability newline )*\\
 \> guest := an ASCII string without newline characters\\
 \> availability := type space price space number\\
 \> type := an ASCII string without newline or space characters\\
 \> price := float encoded as ASCII string\\
 \> number := a long integer encoded as ASCII string
\end{tabbing}

Here, the \texttt{status} field in a response message indicates the
 following:
\begin{tabbing}
\hspace{20pt}\=\kill
 \> 0 - No error, everything is ok\\
 \> 1 - Application error (see result for string description)\\
 \> 2 - Protocol error (usually malformed request)
\end{tabbing}

The \textsc{ASCII} encoding of a float is to be understood as in the
behaviour of the \texttt{Java} method
\texttt{java.lang.Float.toString(float f)}.


\end{document}
